A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-08-26 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: Use existing Peliti citation
As it turns out, there already exists a bibtex entry for Peliti's stat mech book. So I'm using that one instead. It seems more accurate anyway.
NonAdditiveProbability: Add proper citations
Add Grabisch to the bibliography.

We will first review how measure theory is used in classical probability [cite] and classical statistical mechanics \cite{peliti2024}. We will connect this to Choquet theory \cite{phelps2001}, which aims to represent each point of a convex subset as a probability measure over its extreme points (i.e. each ensemble can be represented by a probability measure over pure states). While such representation is always possible, at least in the compact case, it is not unique. Uniqueness requires the space to be a Choquet simplex, the generalization of the finite-dimensional simplex, and represents the classical case. The key insight, then, is that an ensemble space is classical if and only if each ensemble admits a unique decomposition in terms of pure states, it is a Choquet simplex.

@@

We can also generalize a notion of count of states. While in classical statistical mechanics we start with a measure that quantifies the count of states and define the entropy using its logarithm, for the ensemble space we proceed in the opposite way: we start from the entropy and we define the count of states as the exponential of the entropy (i.e. the entropic volume).[cite Hall] From this, we can define a state capacity, which is again a sub-additive measure. It reduces to the Liouville measure in the classical case and the dimensionality of subspaces in the quantum case.

Lastly, we lay down the requirements for a non-additive generalization of probability theory, which should be in principle possible. Since there is a wide literature on the subject, [cite Grabish, Pap, TorraSugeno] we only show that the most used non-additive integrals, the Sugeno and Choquet integral, do not work for our purposes. This means that more work will be needed to develop the appropriate tools.

A final remark before we start. Too often works in the foundations of physics focus on a particular aspect without concern to a broader connection to all relevant mathematical, physical and philosophical aspects, such as information geometry, functional analysis, quantum logic and so on. However, such a complete work would be effectively overwhelming both to produce and to read. The main part of the paper, then, highlights the main conceptual results regarding measure theoretic representations, the appendix goes through the relevant mathematical detail while all the rest is left to our evolving collaborative open source work.[cite our book]

@@

\textbf{Basic requirements for an ensemble space}. In our larger work, we strive to give all axioms and mathematical definitions solid grounding in physical requirements.\footnote{We call this approach Physical Mathematics, as starting from physical ideas recovers the mathematical structures, and only the mathematical structures, that are physically significant.} This way we can develop mathematical frameworks that are physically well founded. Specifically, we are developing a general theory of ensemble spaces, the full treatment of which is beyond the scope of this work. Here we limit ourselves to giving only the initial premises and the final mathematical characterization, with more details in the appendix and in our ongoing work.[cite book]

@@

An ensemble space $\Ens$ will need to satisfy three basic requirements. First, it needs to guarantee experimental verifiability. Mathematically, the ensemble space must be a $\mathsf{T}_0$ second countable topological space, where each open set represents an experimentally verifiable statement[cite book and kevin kelly].\footnote{A verifiable statement is an assertion for which there exists an experimental test that finishes successfully in finite time if and only if the statement is true.} Second, it needs to allow statistical mixtures: if $\ens[a],\ens[b] \in \Ens$ are ensembles, then $p \ens[a] + \bar{p} \ens[b]$ represents the statistical mixture where $\ens[a]$ is taken with probability $p$ and $\ens[b]$ with probability $\bar{p} = 1 - p$. Mathematically, this constrains the ensemble space to be a convex set.\footnote{Only finite mixtures are guaranteed, while the topology will decide whether an infinite mixture $\sum_{i=1}^{\infty} p_i \ens_i$ converges in the space or not.} Third, an entropy function must be defined that characterizes the variability of the elements within an ensemble.

@@

Note that, if not for the lack of additivity over disjoint sets, the fraction capacity would be a probability measure. It is therefore a non-additive probability measure. The name fraction capacity is chosen to both signify the ability of the set $A$ to contain $\ens$ (i.e. fraction capacity of one means $\ens$ is within the convex combinations of $A$) and to indicate that it will be a non-additive monotonic measure, which are called ``capacities'' in some literature.\footnote{ \href{ https://link.springer.com/book/10.1007/978-3-319-03155-2}{cite} }

>>>>

We will first review how measure theory is used in classical probability [cite] and classical statistical mechanics \cite{Peliti2011}. We will connect this to Choquet theory \cite{phelps2001}, which aims to represent each point of a convex subset as a probability measure over its extreme points (i.e. each ensemble can be represented by a probability measure over pure states). While such representation is always possible, at least in the compact case, it is not unique. Uniqueness requires the space to be a Choquet simplex, the generalization of the finite-dimensional simplex, and represents the classical case. The key insight, then, is that an ensemble space is classical if and only if each ensemble admits a unique decomposition in terms of pure states, it is a Choquet simplex.

@@

We can also generalize a notion of count of states. While in classical statistical mechanics we start with a measure that quantifies the count of states and define the entropy using its logarithm, for the ensemble space we proceed in the opposite way: we start from the entropy and we define the count of states as the exponential of the entropy (i.e. the entropic volume)\cite{hall1999}. From this, we can define a state capacity, which is again a sub-additive measure. It reduces to the Liouville measure in the classical case and the dimensionality of subspaces in the quantum case.

Lastly, we lay down the requirements for a non-additive generalization of probability theory, which should be in principle possible. Since there is a wide literature on the subject, \cite{pap2010,pap2013,nonadditive2014,grabisch2016}[cite Grabish, Pap, TorraSugeno] we only show that the most used non-additive integrals, the Sugeno and Choquet integral, do not work for our purposes. This means that more work will be needed to develop the appropriate tools.

A final remark before we start. Too often works in the foundations of physics focus on a particular aspect without concern to a broader connection to all relevant mathematical, physical and philosophical aspects, such as information geometry, functional analysis, quantum logic and so on. However, such a complete work would be effectively overwhelming both to produce and to read. The main part of the paper, then, highlights the main conceptual results regarding measure theoretic representations, the appendix goes through the relevant mathematical detail while all the rest is left to our evolving collaborative open source work.\cite{aop-book}

@@

\textbf{Basic requirements for an ensemble space}. In our larger work, we strive to give all axioms and mathematical definitions solid grounding in physical requirements.\footnote{We call this approach Physical Mathematics, as starting from physical ideas recovers the mathematical structures, and only the mathematical structures, that are physically significant.} This way we can develop mathematical frameworks that are physically well founded. Specifically, we are developing a general theory of ensemble spaces, the full treatment of which is beyond the scope of this work. Here we limit ourselves to giving only the initial premises and the final mathematical characterization, with more details in the appendix and in our ongoing work.\cite{aop-book}

@@

An ensemble space $\Ens$ will need to satisfy three basic requirements. First, it needs to guarantee experimental verifiability. Mathematically, the ensemble space must be a $\mathsf{T}_0$ second countable topological space, where each open set represents an experimentally verifiable statement\cite{aop-book,kelly1996}.\footnote{A verifiable statement is an assertion for which there exists an experimental test that finishes successfully in finite time if and only if the statement is true.} Second, it needs to allow statistical mixtures: if $\ens[a],\ens[b] \in \Ens$ are ensembles, then $p \ens[a] + \bar{p} \ens[b]$ represents the statistical mixture where $\ens[a]$ is taken with probability $p$ and $\ens[b]$ with probability $\bar{p} = 1 - p$. Mathematically, this constrains the ensemble space to be a convex set.\footnote{Only finite mixtures are guaranteed, while the topology will decide whether an infinite mixture $\sum_{i=1}^{\infty} p_i \ens_i$ converges in the space or not.} Third, an entropy function must be defined that characterizes the variability of the elements within an ensemble.

@@

Note that, if not for the lack of additivity over disjoint sets, the fraction capacity would be a probability measure. It is therefore a non-additive probability measure. The name fraction capacity is chosen to both signify the ability of the set $A$ to contain $\ens$ (i.e. fraction capacity of one means $\ens$ is within the convex combinations of $A$) and to indicate that it will be a non-additive monotonic measure, which are called ``capacities'' in some literature.\cite{nonadditive2014}
